---
title: "animatedplyr — Demo"
format:
  html:
    embed-resources: true
    toc: true
---

```{r setup, include=FALSE}
library(animatedplyr)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## The three verbs

### filter()

```{r}
animate_filter(iris, Sepal.Length > 5, seed = 42)
```

### select()

```{r}
animate_select(iris,Sepal.Length, Sepal.Width, seed = 42)
```

### mutate()

```{r}
animate_mutate(mtcars, wt_kg = wt * 453.6, seed = 42)
```

### arrange()

```{r}
animate_arrange(mtcars, desc(mpg), seed = 42)
```

Two sort keys, mixed directions:

```{r}
animate_arrange(palmerpenguins::penguins, species, desc(body_mass_g), seed = 42)
```

### group_by() + summarize(), in one animation

```{r}
animate_group_summarize(mtcars, cyl, n = n(), avg_mpg = mean(mpg), seed = 42)
```

Any summary function `dplyr::summarise()` accepts works — here on the
built-in Kobe Bryant game log:

```{r}
animate_group_summarize(kb_df, Game_Location,
                        n = n(), avg_pts = mean(PTS), sd_pts = sd(PTS),
                        seed = 42)
```

## One-call wrapper

```{r}
with_animation(mtcars, filter(mpg > 19.3), seed = 42)
```

## Reproducibility

Same seed, same sample — render this document twice and these two are identical:

```{r}
animate_filter(mtcars, mpg > 19.3, seed = 7)
```

## Custom configuration

```{r}
big <- animate_config(box_size = 120, font_size = 15, duration = 1800)
animate_select(mtcars, mpg, wt, config = big, seed = 42)
```

## Edge cases become callouts

```{r}
animate_filter(mtcars, mpg > 0)      # all rows pass
```

```{r}
animate_filter(mtcars, mpg > 1000)   # no rows pass
```

```{r}
animate_arrange(data.frame(x = 1:8), x)   # rows already in order
```

## Disclosure pill

`mtcars` has 32 rows and 11 columns; the animation samples 5 rows and at most
4 columns, and the gray pill below the grid says what was hidden.
