---
title: "Animating dplyr Verbs with NBA GOAT Data"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Animating dplyr Verbs with NBA GOAT Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  comment   = "#>"
)
```

## 1. Introduction

### The `animatedplyr` Package

`animatedplyr` generates step-by-step HTML animations that show how the core **dplyr data-wrangling verbs** transform a data frame:

| Verb                       | What it does                   | Animator                    |
|----------------------------|--------------------------------|-----------------------------|
| `select()`                 | Keep or drop columns           | `animate_select()`          |
| `filter()`                 | Keep rows matching a condition | `animate_filter()`          |
| `mutate()`                 | Add or modify columns          | `animate_mutate()`          |
| `arrange()`                | Reorder rows                   | `animate_arrange()`         |
| `group_by()` + `summarize()` | Aggregate one row per group  | `animate_group_summarize()` |

Each animation renders inline below a code chunk in Quarto / R Markdown and in the RStudio / Positron Viewer pane at the console. Because big data frames are hard to follow on screen, `animatedplyr` applies *smart sampling*: it shows at most **5 rows and 4 columns** while always including the columns that matter for the operation being animated.

### The NBA GOATs Dataset

`kb_df` contains a **game-by-game log** of Kobe Bryant's first 15 NBA regular seasons (1996–2011). Each of the **1,156 rows** is one game; the **38 columns** record shooting stats, rebounding, playmaking, game context, and computed indicators like double-doubles and the Hollinger Game Score.

This dataset is ideal for learning data wrangling because:

-   It has a mix of numeric, character, and date columns.
-   It rewards filtering (high-scoring games, road wins, clutch performances).
-   It rewards selecting (focusing on just the stat columns you care about).
-   It rewards mutating (building efficiency metrics, classifying performances).

------------------------------------------------------------------------

## 2. Loading the Dataset

```{r load-pkg}
library(animatedplyr)
library(dplyr)
```

The `kb_df` data frame is bundled with the package and loads lazily — just name it and R finds it automatically.

```{r load-data}
data(kb_df)   # explicit load (optional but good practice in scripts)
dim(kb_df)
```

### Quick peek

```{r glimpse}
dplyr::glimpse(kb_df)
```

### Key columns at a glance

```{r head-key}
head(kb_df[, c("Season", "Date", "Game_Location",
                    "Game_Outcome", "PTS", "AST", "TRB", "GmSc")], 8)
```

### Summary statistics

```{r summary-pts}
summary(kb_df[, c("PTS", "AST", "TRB", "MP", "GmSc")])
```

Kobe's highest single-game scoring output:

```{r max-pts}
kb_df[which.max(kb_df$PTS),
          c("Date", "Season", "Opp", "Game_Location", "PTS", "FGA", "GmSc")]
```

------------------------------------------------------------------------

## 3. `select` — Choosing Columns

`animate_select()` shows which columns survive and which fade away. The animation plays two steps:

1.  **Before** — the full set of sampled columns is displayed.
2.  **After** — dropped columns fade out; surviving columns recenter.

### 3.1 Select a handful of columns by name

The most common use case: pick exactly the columns you need.

```{r select-basic}
animate_select(kb_df, Name, Season, TRB, seed = 1)
```

### 3.2 Select game-context columns

Narrow the view to the situational columns that describe *where* and *how* a game was played.

```{r select-context}
animate_select(kb_df,
               Date, Season, Game_Location, Game_Outcome, Point_Margin,
               seed = 7)
```

### 3.3 Select shooting statistics only

```{r select-shooting}
animate_select(kb_df, PTS, FG, FGA, FG_Percent, FT, FTA, FT_Percent,
               seed = 42)
```

### 3.4 Using the `with_animation()` wrapper

`with_animation()` lets you write the exact `dplyr` expression you already know, and the package routes it to the right animator automatically.

```{r select-with-animation}
with_animation(kb_df,
               select(Name, Season, PTS, AST, TRB, BLK, STL),
               seed = 10)
```

### 3.5 Customising the animation appearance

`animate_config()` lets you slow the animation down and increase the font size — handy in a classroom projector setting.

```{r select-config}
cfg <- animate_config(duration = 1800, font_size = 15)

animate_select(kb_df, Season, Tm, Opp, PTS, AST, TRB,
               config = cfg, seed = 5)
```

------------------------------------------------------------------------

## 4. `filter` — Subsetting Rows

`animate_filter()` shows the condition being evaluated row by row:

1.  **Before** — all sampled rows are displayed; kept rows turn **green**, dropped rows turn **red**.
2.  **After** — the red rows collapse away, leaving only the kept rows.

### 4.1 Filter for home games

```{r filter-home}
animate_filter(kb_df, Game_Location == "Home", seed = 1)
```

### 4.2 Filter for wins on the road

Combining two conditions with `&` works exactly as in `dplyr`.

```{r filter-road-wins}
animate_filter(kb_df,
               Game_Location == "Away" & Game_Outcome == "W",
               seed = 3)
```

### 4.3 Filter for elite scoring performances (40+ points)

```{r filter-40plus}
animate_filter(kb_df, PTS >= 40, seed = 42)
```

Let's also see how many games that is:

```{r count-40plus}
sum(kb_df$PTS >= 40, na.rm = TRUE)
```

### 4.4 Filter for the iconic 81-point game

```{r filter-81}
animate_filter(kb_df, PTS == 81, seed = 1)
```

### 4.5 Filter for double-doubles

```{r filter-double-double}
animate_filter(kb_df, DD == 1, seed = 2)
```

### 4.6 Filter for efficient performances (Game Score above 30)

The **Game Score** (column `GmSc`) is Hollinger's single-number efficiency summary. A score above 30 is an elite performance.

```{r filter-gmscore}
animate_filter(kb_df, GmSc >= 30, seed = 6)
```

### 4.7 Using the `with_animation()` wrapper for filter

```{r filter-with-animation}
with_animation(kb_df,
               filter(PTS > 30 & Game_Outcome == "W"),
               seed = 9)
```

### 4.8 Season-specific filtering

```{r filter-season-8}
animate_filter(kb_df, Season == "season_8", seed = 5)
```

------------------------------------------------------------------------

## 5. `mutate` — Creating New Variables

`animate_mutate()` shows how a new column is computed:

1.  **Before** — the source column(s) the expression uses are highlighted in the animation.
2.  **After** — the new column pops into view with its computed values.

> **Tip:** `animate_mutate()` animates one new column at a time. Chain calls for multiple mutations.

### 5.1 Compute field-goal efficiency rate

Raw points are important, but *how efficiently* were they scored? Let's compute the true shooting percentage (a simplified version: `PTS / FGA`).

```{r mutate-efficiency}
animate_mutate(kb_df, pts_per_attempt = PTS / FGA, seed = 1)
```

### 5.2 Classify games as high-scoring (30+ points)

```{r mutate-high-score}
animate_mutate(kb_df, high_multi = PTS * 30, seed = 4)
```

### 5.3 Create a home/away numeric flag

Convert the character column `Game_Location` into a binary `is_home` indicator.

```{r mutate-home-flag}
animate_mutate(kb_df, is_home = Game_Location == "Home", seed = 7)
```

### 5.4 Compute points above average

This shows how a value-relative-to-mean transformation looks step by step.

```{r mutate-above-avg}
avg_pts <- mean(kb_df$PTS, na.rm = TRUE)
animate_mutate(kb_df, pts_above_avg = PTS - avg_pts, seed = 3)
```

### 5.5 Compute a combined scoring + playmaking metric

```{r mutate-impact}
animate_mutate(kb_df, scoring_playmaking = PTS + (AST * 2), seed = 2)
```

### 5.6 Label the game result as a descriptive string

```{r mutate-result-label}
animate_mutate(kb_df,
               result_label = ifelse(Game_Outcome == "W", "Win", "Loss"),
               seed = 5)
```

### 5.7 Using the `with_animation()` wrapper for mutate

```{r mutate-with-animation}
with_animation(kb_df,
               mutate(blk_stl_sum = BLK + STL),
               seed = 8)
```

### 5.8 Chaining animations to show a multi-step pipeline

Below we build up a transformation step-by-step — first selecting the columns we care about, then filtering for winning performances, then mutating to add an efficiency metric — with a separate animation for each step.

```{r pipeline-select}
# Step 1 — narrow the view
animate_select(kb_df, Season, PTS, AST, TRB, FGA, Game_Outcome, seed = 1)
```

```{r pipeline-filter}
# Step 2 — keep only wins
animate_filter(kb_df, Game_Outcome == "W", seed = 1)
```

```{r pipeline-mutate}
# Step 3 — add an efficiency column
animate_mutate(kb_df, pts_per_fg_attempt = PTS / FGA, seed = 1)
```

------------------------------------------------------------------------

## 6. `arrange` — Reordering Rows

`animate_arrange()` shows how sorting works without touching the data itself:

1.  **Before** — the sampled rows in their current order.
2.  **Sort key** — the column(s) driving the sort light up, with the direction named.
3.  **Reshuffle** — the rows visibly re-order.
4.  **After** — the closing label makes the key point: *rows reordered, none added or removed*. Same rows, same dimensions — `arrange()` changes ORDER, never CONTENT.

The sampler deliberately picks rows that are **not** already in sorted order, so every animation has a visible reshuffle.

### 6.1 Best scoring games first

```{r arrange-pts}
animate_arrange(kb_df, desc(PTS), seed = 1)
```

### 6.2 Two sort keys, mixed directions

Sort by season, and within each season put the most efficient games
(highest Game Score) first — exactly the `dplyr` semantics, ties and all.

```{r arrange-season-gmsc}
animate_arrange(kb_df, Season, desc(GmSc), seed = 4)
```

### 6.3 Using the `with_animation()` wrapper for arrange

```{r arrange-with-animation}
with_animation(kb_df, arrange(desc(TRB)), seed = 6)
```

### 6.4 When the rows are already in order

`kb_df` is a chronological game log, so sorting by `Date` changes nothing —
and the package says so with a callout instead of a pointless animation.

```{r arrange-sorted-callout}
animate_arrange(kb_df, Date, seed = 1)
```

------------------------------------------------------------------------

## 7. `group_by` + `summarize` — Aggregating by Group

`animate_group_summarize()` animates the whole aggregation story in one call
(no chaining needed):

1.  **Before** — raw sampled rows, ungrouped.
2.  **group_by** — rows gather into dashed, color-coded boxes, one per group, each with a badge like "Home × 3". The label stresses that the data is *unchanged* — `group_by()` only tags rows.
3.  **Gather** — within each box, rows collapse toward one.
4.  **summarize** — one row per group pops in with the computed values, outlined in its group's color so you can trace where it came from.

Up to 3 groups are shown with deliberately **uneven** sizes, so students see that groups differ before they collapse; the badge count always equals `n()`.

### 7.1 Home vs. away scoring

```{r group-location}
animate_group_summarize(kb_df, Game_Location,
                        n = n(), avg_pts = mean(PTS),
                        seed = 1)
```

### 7.2 Wins vs. losses — any summary function works

Anything `dplyr::summarise()` accepts is fair game: `n()`, `mean()`, `sd()`,
`median()`, `max()`, ...

```{r group-outcome}
animate_group_summarize(kb_df, Game_Outcome,
                        avg_margin = mean(Point_Margin),
                        best_game  = max(PTS),
                        seed = 2)
```

### 7.3 Many groups? The pill tells you what's hidden

`kb_df` spans 15 seasons; the animation shows the 3 most frequent groups and
the gray pill discloses the rest ("+12 groups not shown").

```{r group-season}
animate_group_summarize(kb_df, Season,
                        avg_gmsc = mean(GmSc),
                        seed = 3)
```

### 7.4 British spelling included

`animate_group_summarise()` is an exact alias — the title chip follows your
spelling.

```{r group-summarise-alias}
animate_group_summarise(kb_df, Game_Location,
                        median_min = median(MP),
                        seed = 5)
```

------------------------------------------------------------------------

## Further Reading

-   `?animate_select`, `?animate_filter`, `?animate_mutate`,
    `?animate_arrange`, `?animate_group_summarize` — per-function docs.
-   `?animate_config` — tuning box size, font, animation speed, and column colors.
-   `?with_animation` — the single-call wrapper for classroom use.
-   `?kb_df`, `?lb_df`, `?mj_df` — player game-log dataset docs.
-   `?variable_names_df` — plain-English descriptions of every column.
-   `?divisions_df` — team abbreviation to division/conference lookup.